Speeding up decimal multiplication

### 2020-11-27 paper/paper.tex
Update

\title{Speeding up decimal multiplication}
\author{Viktor Krapivensky}
\date{%
    \today
}

@@

It is a common theme in practical analysis of algorithms that ``fancy'' algorithms are slow when $n$ is small.
For example, virtually all partical implementations of the ``fancy'' sorting algorithms that are divide-and-conquer in nature --- quick sort and merge sort --- do in fact fall back to ``dumb'' quadratic sorts if the size of the input if less than some threshold.
In some cases, it is not even a matter of ``fancy'' versus ``dumb'' dichotomy, but rather there is a ``hierarchy of fanciness'':
a group of algorithms with different asymptotic behavior such that it makes sense to use each one of them only for some range of values of $n.$
To give an example, \textbf{libgmp} employs the following ``hierarchy of fanciness'' for multiplication~\cite{gmp_fh}:
first goes the ``basecase'' (quadratic) algorithm; then, the Karatsuba algorithm; then, variations of the Toom-Cook algorithm (Toom-3; then Toom-4; then Toom-6.5; then Toom-8.5); finally, the Schönhage--Strassen algorithm (which is Fourier transform-based).

@@

The \textbf{mpdecimal} library uses yet another method of modular reduction on 64-bit systems;
it is based on the use of prime moduli of form $2^{64} - 2^{n} + 1.$
Generally, such primes are knows as Solinas primes, or generalized Mersenne
primes~\cite{Solinas},
defined as primes of form $f(2^m),$ where $f(x)$ is a low-degree polynomial with small integer
coefficients.

@@

Note that it uses an unspecified algorithm to perform primality testing.
We use the Miller-Rabin primality test with the first 12 prime numbers as bases; it has been proven
in \cite{primebases} that, for values less than $2^{64},$ this is enough to guarantee correctness.

@@

For the four-step algorithm, this means no bit-reversal permutation or matrix transposition is ever
needed: we can also simultaneously omit the bit-reversal permutation steps in DIF and DIT
if we use DIF for the direct transforms in definition of four-step DFT, and DIT for inverse
transforms in the definition of four-step IDFT.

@@

We provide the pseudocode for case of $C=2.$
Note it uses the routine $\mathrm{TransposeSquareSubMatrix}$.
It does not have to be implemented exactly in a way we demonstrated above;
it just needs to have the same semantics.
It can be implemented in a cache-friendly way by dividing the square submatrix into
tiles, perhaps recursively, transposing the elements inside tiles, and then swapping the tiles
themselves; see~\cite{cache_square_mat_trans} for details.

@@

Note that we use this optimization much more sparingly than we theoretically could, because
allowing non-local propagation of ``unadjustment'' would be a nightmare from the standpoint of
reliability and debuggability: we would need to keep track of what can, and what can not, be
``unadjusted'' at each step.

@@

Since both rely on Cooley-Tukey,
the graphs of time as a function of input size $n$ (in decimal digits)
would be staircase-like.
Define ``threshold input size'' for some fixed implementation as any value of $n$ where the
discontinuity occurs --- a new ``stair'' pops up.

>>>>

\title{Speeding up decimal multiplication}
\author{Viktor Krapivensky}
\date{%
    November 25, 2020
}

@@

It is a common theme in practical analysis of algorithms that ``fancy'' algorithms are slow when $n$ is small.
For example, virtually all partical implementations of the ``fancy'' sorting algorithms that are divide-and-conquer in nature --- quick sort and merge sort --- do in fact fall back to ``dumb'' quadratic sorts if the size of the input is less than some threshold.
In some cases, it is not even a matter of ``fancy'' versus ``dumb'' dichotomy, but rather there is a ``hierarchy of fanciness'':
a group of algorithms with different asymptotic behavior such that it makes sense to use each one of them only for some range of values of $n.$
To give an example, \textbf{libgmp} employs the following ``hierarchy of fanciness'' for multiplication~\cite{gmp_fh}:
first goes the ``basecase'' (quadratic) algorithm; then, the Karatsuba algorithm; then, variations of the Toom-Cook algorithm (Toom-3; then Toom-4; then Toom-6.5; then Toom-8.5); finally, the Schönhage--Strassen algorithm (which is Fourier transform-based).

@@

The \textbf{mpdecimal} library uses yet another method of modular reduction on 64-bit systems;
it is based on the use of prime moduli of form $2^{64} - 2^{n} + 1.$
Generally, such primes are known as Solinas primes, or generalized Mersenne
primes~\cite{Solinas},
defined as primes of form $f(2^m),$ where $f(x)$ is a low-degree polynomial with small integer
coefficients.

@@

Note that it uses an unspecified algorithm to perform primality testing.
We use the Miller-Rabin primality test with the first 12 prime numbers as bases; it has been proven
in~\cite{primebases} that, for values less than $2^{64},$ this is enough to guarantee correctness.

@@

For the four-step algorithm, this means no bit-reversal permutation or matrix transposition is ever
needed: we can also simultaneously omit the bit-reversal permutation steps in DIF and DIT
if we use DIF for the direct transforms in definition of four-step DFT, and DIT for inverse
transforms in the definition of four-step IDFT*.

@@

We provide the pseudocode for the case of $C=2.$
Note it uses the routine $\mathrm{TransposeSquareSubMatrix}$.
It does not have to be implemented exactly in a way we demonstrated above;
it just needs to have the same semantics.
It can be implemented in a cache-friendly way by dividing the square submatrix into
tiles, perhaps recursively, transposing the elements inside tiles, and then swapping the tiles
themselves; see~\cite{cache_square_mat_trans} for details.

@@

Note that we use this optimization much more sparingly than we theoretically could, because
allowing non-local propagation of ``unadjustment'' would be a nightmare from the standpoints of
reliability and debuggability: we would need to keep track of what can, and what can not, be
``unadjusted'' at each step.

@@

Since both rely on Cooley-Tukey,
the graphs of time as a function of input size $n$ (in decimal digits)
would be staircase-like.
Define ``threshold input size'' for some fixed implementation as any value of $n$ where a
discontinuity occurs --- a new ``stair'' pops up.
